Ricci-Ollivier Curvature of the Rooted Phylogenetic Subtree-Prune-Regraft Graph

### 2015-07-07 curvature.tex
minor edits

We required several new computational tools. 
We present a fast new algorithm for computing rSPR graphs from a set of trees, reducing the time to do so from $\OhOf{m^2n}$ to $\OhOf{mn^3}$ for a set of $m$ trees with $n$ leaves.
As the full rSPR graph on trees with $n$ leaves contains $(2n-3)!!$ trees, this is a significant improvement in practice for exploring large subsets of the graph (or, as we do here, the full graph for small numbers of leaves).
By exploiting symmetries in the rSPR graph, we were able to calculate all of the curvatures for pairs of trees with up to seven leaves.
By carefully examining the overlap in rSPR moves, we present a new method for computing the degree of a tree in the rSPR graph that allows one to select an rSPR neighbor uniformly at random in linear-time without explicitly generating the graph.
This is a major improvement over the sampling methods used in current software programs such as MrBayes which do not sample SPR moves uniformly.

@@

In previous work~\cite{Whidden2015-yi}, we constructed (unrooted) SPR graphs from subsets of $m$ high probability trees sampled from phylogenetic posteriors to compare mixing and identify graph effects.
Although the SPR distance (rooted and unrooted) is NP-hard to compute~\cite{bordewich05,hickey2008sdc}, it is fixed-parameter tractable with respect to the distance in the rooted case~\cite{bordewich05}.
In particular, one can determine in $\OhOf{n}$-time whether two rooted phylogenetic trees are adjacent in the rSPR graph ($\OhOf{n^2}$-time for unrooted trees) using the algorithms of Whidden et al.~\cite{whidden2009unifying,whidden2010fast, whidden2013hybridization,Whidden2015-yi}.
We applied this method comparing each of the $m$ trees pairwise to identify adjacencies, requiring a total of $\OhOf{m^2n}$-time ($\OhOf{m^2 n^2}$-time in the unrooted case).
However, this method is infeasible when applied to construct graphs with 7 or more leaves, due to the rapidly growing $\OhOf{m^2}$ factor.

@@

We now bound the difference between the curvature of the uniform random walk $\curvature{T,S}$ and that of the Metropolis-Hastings (MH) random walk $\curvature{\MH; T,S}$.
Recall that this random walk proposes a move from a tree $T$ to a neighbor tree $S$ uniformly at random and then accepts the move according to the Hastings ratio, \cuttable{which in this case is }$\min\left(1, \frac{\degree{T}}{\degree{S}}\right)$.
The mass distribution \cuttable{for the MH random walk }thus leaves a portion of mass at the origin tree, proportional to the relative degree difference of its higher degree neighbors.
Note that \cuttable{the same statement and proof of }Lemma~\ref{lem:asymptotic} holds with $\curvature{T,S}$ and $\ric{T,S}$ replaced by the MH curvatures $\curvature{\MH; T,S}$ and $\ric{\MH; T,S}$, respectively.

>>>>

We required several new computational tools. 
We present a fast new algorithm for computing rSPR graphs from a set of trees, reducing the time to do so from $\OhOf{m^2n}$ to $\OhOf{mn^3}$ for a set of $m$ trees with $n$ leaves.
As the full rSPR graph on trees with $n$ leaves contains $(2n-3)!!$ trees, this is a significant improvement in practice for exploring large subsets of the graph (or, as we do here, the full graph for small numbers of leaves).
By exploiting symmetries in the rSPR graph, we were able to calculate all of the curvatures for pairs of trees with up to seven leaves.
By carefully examining the overlap in rSPR moves, we present a new method for computing the degree of a tree in the rSPR graph that allows one to select an rSPR neighbor uniformly at random in linear-time without explicitly generating the graph.
This is a major improvement over the sampling methods used in current software programs such as MrBayes which do not propose SPR moves uniformly.

@@

In previous work~\cite{Whidden2015-yi}, we constructed (unrooted) SPR graphs from subsets of $m$ high probability trees sampled from phylogenetic posteriors to compare mixing and identify graph effects.
Although the SPR distance (rooted and unrooted) is NP-hard to compute~\cite{bordewich05,hickey2008sdc}, it is fixed-parameter tractable with respect to the distance in the rooted case~\cite{bordewich05}.
In particular, one can determine in $\OhOf{n}$-time whether two rooted phylogenetic trees are adjacent in the rSPR graph ($\OhOf{n^2}$-time for unrooted trees) using the algorithms of Whidden et al.~\cite{whidden2009unifying,whidden2010fast, whidden2013hybridization,Whidden2015-yi}.
We applied this method comparing each of the $m$ trees pairwise to identify adjacencies, requiring a total of $\OhOf{m^2n}$-time ($\OhOf{m^2 n^2}$-time in the unrooted case).
However, this method is impractical when applied to construct graphs with 7 or more leaves, due to the rapidly growing $\OhOf{m^2}$ factor.

@@

Finally, we bound the difference between the curvature of the uniform random walk $\curvature{T,S}$ and that of the Metropolis-Hastings (MH) random walk $\curvature{\MH; T,S}$.
Recall that this random walk proposes a move from a tree $T$ to a neighbor tree $S$ uniformly at random and then accepts the move according to the Hastings ratio, \cuttable{which in this case is }$\min\left(1, \frac{\degree{T}}{\degree{S}}\right)$.
The mass distribution \cuttable{for the MH random walk }thus leaves a portion of mass at the origin tree, proportional to the relative degree difference of its higher degree neighbors.
Note that \cuttable{the same statement and proof of }Lemma~\ref{lem:asymptotic} holds with $\curvature{T,S}$ and $\ric{T,S}$ replaced by the MH curvatures $\curvature{\MH; T,S}$ and $\ric{\MH; T,S}$, respectively.
